Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia

### 2024-11-05 analysis/paper/paper.qmd
- corrected refs - grammar corrections

Several key topics in the study of hunter-gatherer behaviour and organization have been explored in different geographic regions and chronological periods through raw material analysis, such as modalities of procurement and mobility strategies [@ambroseSocialEcologicalModels1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccallBehavioralEcologicalModels2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallonSocialNetworksInformation2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999]. These analyses are often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins. The ultimate aim is to identify the corresponding sources, enabling, for example, a precise differentiation between local and non-local raw materials, and examine their distribution within a site over different occupations and periods.

@@

Both these areas have been previously interpreted as seasonal residential camps, repeatedly used for extended stays, due to the abundance of lithic debitage, stone tools, heat-cracked rocks related to grease rendering activities, large quantities of faunal remains (both marine and terrestrial) and the presence of ornaments and portable art [@manneIntensiveSubsistencePractices2012]. The analysis of lithic assemblages and retouched frequency of lithic assemblages from the Terrace and Shelter areas also corroborate this interpretation [@cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009; @cascalheira_2017]. Cascalheira et al. [-@cascalheira_2017] show that, with the exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter, all other UP occupations are composed of high-density assemblages with low degree of retouch, correspondent to an residential base-camp occupation, for extended periods of time.

@@

The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources. Two of the main geological units which characterize the Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the South Portuguese Zone [@fernandesNewEvidenceConcerning2012; @oliveira_carta_1984-1; @onezimeNewGeodynamicInterpretation2003], located in the north sector of the region, and characterized by schists, greywackes and quartzites; and 2) the Algarve basin in the south, which is characterized by limestones, dolomites and marls [@terrinha_bacia_2006]. It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules/beds can be found [@manuppellaNoticiaExplicativaFolha1987; @manuppellaNoticiaExplicativaFolha2007; @ribeiro_evolucao_2005; @rocha_noticia_1979; @rocha_noticia_1983; @terrinha_bacia_2006].

@@

We employed both macroscopic and petrographic methods to achieve a more comprehensive characterization of the materials, aiming to offset the inherent limitations of each method [@luedtke_archaeologists_1992].

Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustilloMacroscopicClassificationFlint2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=4458). The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins. Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustilloMacroscopicClassificationFlint2009; @tarrino_materias_2013]. Therefore, we implemented an initial phase of macroscopic analysis, subsequently augmented by petrographic study to diminish the subjectivity of the analysis and enhance the material characterization with petrographic data. The synergistic application of these two methods has been previously validated in the characterization of Algarve region cherts, following recent geological survey efforts [@belmiroCreatingFramesReference2023]. Hence, an extensive macroscopic and petrographic collection was already available for comparative analysis with archaeological materials, employing consistent methodologies. This reference collection is the Lusolit lithotheque, hosted physically at ICArEHB (University of Algarve, Faro) and digitally at www.lusolit.icarehb.com, and aided in the characterization and source attribution of cherts.

@@

The macroscopic analysis of the archaeological materials was divided into two steps. The first step included a preliminary characterization using a hand lens of 10x magnification. This step allowed us to create types and subtypes based on similar macroscopic characteristics frequently acknowledged to be useful to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1). In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types - the data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1). Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis was simplified and focused on collecting data related to weight, cortex, and alterations. This first step was applied to a total of 4458 artefacts: a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 7 to 4) and 831 chert artefacts from the Solutrean sequence of the Shelter area (layers A, B, C; all samples without an attributed layer on the database were excluded from the analysis).

@@

```{r}
#| label: fig-local-types
#| fig-cap: Variability of local chert types and main petrographic characteristics. a) Type 1 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope of white patina and fractures; a2) microscopic view of thin section (XPL), with visible fibrous chalcedony composition and without allochems. b) Type 3 macroscopic and microscopic views: b1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; b2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides. c-d) Type 2 macroscopic and microscopic views of multicoloured (c) and single-colour (d) varieties. c1) macroscopic view with a stereomicroscope of the multicoloured bands with iron oxides and unidentifiable fossils; d1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; c2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; d2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides. e-f) Type 4 macroscopic and microscopic views: e) macroscopic view with a stereomicroscope of unidentifiable fossils and white splotches; f1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils within white and yellow splotches; f2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides; f3) microscopic view of thin section (PPL) with a concentration of fossils (possibly sponge spicules) and a fracture. g-h) Type 5 macroscopic and microscopic views: g) macroscopic view with a stereomicroscope; h1) macroscopic view with a stereomicroscope with visible unidentifiable fossils; h2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; h3) microscopic view of thin section (XPL) with small unidentifiable fossils, iron oxides and a possible ostracod replaced by fibrous chalcedony and macrocrystalline quartz.

@@

```{r}
#| label: fig-distance-types
#| fig-cap: Variability of non-local types and main petrographic characteristics. a-b) Type 6 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope with a small degree of white patina and no visible fossils; b1) macroscopic view with a stereomicroscope of a thick alteration rind but translucent interior where some unidentifiable fossils are visible; a2) microscopic view of thin section (PPL) with a mudstone texture; b2) microscopic of thin section (XPL) with small unidentifiable fossils and a laminar microstructure. c-d) Type 7 macroscopic and microscopic views of the peloidal facies (c) and oolitic facies (d): c1) macroscopic view with a stereomicroscope of peloids and sponge spicules; d1) macroscopic view with a stereomicroscope of poorly preserved ooids; c2) microscopic view of thin section (PPL) with packstone texture composed mainly of peloids; d2) microscopic view of thin section (PPL) with packstone texture composed mainly of ooids. e-f) Type 8 macroscopic and microscopic views, with foraminifera represented by the white and black arrows: e) macroscopic view with a stereomicroscope with abundant unidentifiable fossils and foraminifera; f1) macroscopic view with a stereomicroscope with abundant unidentifiable fossils but no visible foraminifera; f2) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which are not visible macroscopically; f3) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which not are visible macroscopically. g) Type 9 macroscopic and microscopic views: g1) macroscopic view with a stereomicroscope of unidentifiable fossils and two different fabrics; g2) microscopic view of thin section (XPL) with packstone texture composed of concentrations of unidentifiable fossils and opaques which create a banded microstructure. h) Type 10 macroscopic and microscopic views: h1) macroscopic view with a stereomicroscope of ooids and fractures filled with chalcedony; h2) microscopic view of thin section (PPL) with packstone texture composed of ooids with different levels of preservation.

@@

Type 10 (@fig-distance-types h) is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids. It represents \~1% of the overall Terrace cherts and less than 0.1% of the chert in the Shelter. Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: brown (opaque) and black (sub-translucent) fabric (@fig-distance-types h1). The sample chosen for the thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the fabrics present in the artefacts. Allochems are common iron oxides and very frequent ooids. Bioclasts are rare, poorly preserved, and replaced by quartz. The porosity is intraparticle and vug (10%). The ooids are distributed homogeneously across the sample (although macroscopically visible in the brown fabric) and are highly abundant and concentrated. They are poorly sorted, varying between 500 and 20 μm (@fig-distance-types h2), show round-to-elliptical shapes and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010]. This chert shows no similarities to any consulted reference material. Despite the existence of oolitic limestones in the Algarve region correspondent to Middle and Upper Jurassic formations and often described as whiteish (Rocha et al., 1979), the presence of oolitic chert in the region is not recognized in geological and archaeological literature [@cardosoSobrePresencaLaminas2018; @rocha_noticia_1979; @verissimo_jazidas_2004] and the visit of outcrops mentioned in the geological cartography did not allow their identification.

@@

```{r}
#| label: fig-small-types
#| fig-cap: Macroscopic view with a stereomicroscope of non-local Type 11 and TL cherts. Type 11 – detail of banded patterns composed of parallel different fabrics; TL01 – detail of the jasper and fractures filled with chalcedony; TL02 – detail of the chalcedony with unidentified orange inclusions; TL03 – detail of two round inclusions, possibly iron oxides; TL04 - detail of massive chert with fractures filled with chalcedony; TL05 – detail of the banded pattern within the grey mudstone texture; TL06 - detail of the small, finely sorted ooids; TL07 – detail of the dendritic laminations, possibly composed of iron oxides; TL08 – detail of the mudstone structure, without the presence of allochems; TL09a/b – detail of the mossy, orange and red inclusions within the chalcedony; TL10 – detail of the speckling and laminated patterns within the mudstone structure; TL11 – detail of the banding within the mudstone structure; TL12 – detail of the poorly preserved ooids, only visible under the stereomicroscope; TL13 – detail of the bands and laminations within the mudstone texture fabric; TL14 – detail of an unidentified benthic foraminifera; TL15 – detail of the interface between the massive brown fabric and splotches; TL16 - detail of the of the massive structure with unidentifiable inclusions.

@@

Jasper is known in the South Portuguese Zone area [@oliveira_carta_1984-1], from Southern/Central Portugal to Spain, north of the Algarve (from Aljustrel to Beja; @fig-map), suggesting that the source of TL01 may be from south/central Portugal at a minimum distance of 100 km north and northeast from Vale Boi (as the crow flies).

@@

In the Gravettian occupations (levels 6 and 7) and at the start of the Proto-Solutrean (level 5) local cherts make up \~45-60%, while the combined non-local cherts make up \~65-40%. Level 6 shows the highest percentages of all non-local chert use, with specific types like Type 6 representing \~30%, although when considering the sum of weight, the percentages reduce, and non-local chert represents \~50% of total chert. The percentages of local chert increase in the following occupations. In the Proto-Solutrean (levels 4E and 4E/5) and Solutrean (levels 4, 4C, and 4D) local cherts make up more than \~70% of the sample and 80% when considering the sum of weight, while non-local cherts decrease significantly and represent percentages lower than 30%. Despite the increase of local cherts during the Proto-Solutrean and Solutrean, there are still statistically significant differences between the local types present in the two technocomplexes. During the top Proto-Solutrean occupations the grey/white cherts (Type 4) are the most present chert with percentages between 35-20%, while in the Solutrean levels, the yellow/brown cherts (Type 2) are present in the highest percentage, at \~30%.

```{r}
#| label: fig-type-per-level
#| fig-cap: Terrace area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples. Only percentages superior to the 5% threshold are shown in the figure. Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts. Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors). Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors).
#| fig-width: 10

@@

When plotting the weight, it is apparent that most of the samples are located under the 10 gr mark, with median and bottom quartiles often falling under the 5 gr line (@fig-weight). Local types, such as Type 2, Type 4, and Type 5, although mostly represented by artefacts with a small mass (\~3 gr), show the heaviest and highest number of heavy artefacts. In comparison, other types, although showing some variation in the weight of artefacts, show smaller artefacts and their means and medians are frequently closer to the bottom quartile. As seen in Figure 3 B, non-local types show larger weights from level 5 to level 7.

```{r}
#| label: fig-weight
#| fig-cap: "Terrace boxplots of chert type weight (gr) per level. Diamond shape within the boxplots represents weight mean. Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts. Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors). Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors)."
#| #| fig-width: 10

@@

```{r}
#| label: fig-analysis_shelter
#| fig-cap: "a-b) Shelter area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples. Only percentages superior to the 5% threshold are shown in the figure. c): Shelter boxplots of chert type weight (gr) per level. Type 1 to 5 (yellow and red colors) represent local chalcedony and cherts. Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colors). Type 8, 10 and 11 represent non-local cherts with non-identified provinience (blue and green colors)."
#| #| fig-width: 10

@@

A similar case is observed for the jaspers located in the SPZ [@oliveira_carta_1984-1], located at a minimum of 100 km distance from Vale Boi, as the crow flies. Despite being a source located at closer distances when compared to the cherts of Central Portugal or from Southwestern Spain, the jasper occurs in trace amounts (n=3) and only in the Gravettian layers (layer 6). Occurring North and Northeast, the limited amount of jasper in the assemblage may either represent a limited use of these resources in detriment of the cherts from the Lisbon and Rio Maior area, or suggest that they were used and knapped elsewhere, but not transported, either by the same groups or through exchanges by different groups.

>>>>

Several key topics in the study of hunter-gatherer behaviour and organisation have been explored in different geographic regions and chronological periods through raw material analysis, such as modalities of procurement and mobility strategies [@ambroseSocialEcologicalModels1990; @binfordOrganizationFormationProcesses1979; @binford_righteous_1985; @gouldEmpiricistStrikesBack1985; @gouldLithicProcurementCentral1985; @kuhnUnpackingReductionLithic1991; @mccallBehavioralEcologicalModels2007], occupation types and their duration [@kuhnUpperPaleolithicRaw2004; @surovellBehavioralEcologyLithic2009], the establishment and dimension of social networks [@whallonSocialNetworksInformation2006], as well as exchanges between groups or individuals [@gamblePalaeolithicSocietiesEurope1999]. These analyses are often achieved by systematically characterizing geological and archaeological raw materials and establishing correlations between samples from both origins. The ultimate aim is to identify the corresponding sources, enabling, for example, a precise differentiation between local and non-local raw materials, and examine their distribution within a site over different occupations and periods.

@@

Both these areas have been previously interpreted as seasonal residential camps, repeatedly used for extended stays, due to the abundance of lithic debitage, stone tools, heat-cracked rocks related to grease rendering activities, large quantities of faunal remains (both marine and terrestrial) and the presence of ornaments and portable art [@manneIntensiveSubsistencePractices2012]. The analysis of lithic assemblages and retouched frequency of lithic assemblages from the Terrace and Shelter areas also corroborate this interpretation [@cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009; @cascalheira_2017]. Cascalheira et al. [-@cascalheira_2017] show that, with the exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter, all other UP occupations are composed of high-density assemblages with low degree of retouch, correspondent to a residential base-camp occupation, for extended periods of time.

@@

The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources. Two of the main geological units which characterize the Algarve are: 1) the Baixo Alentejo Flyschoid Group, part of the South Portuguese Zone [@fernandesNewEvidenceConcerning2012; @oliveiraCartaGeologicaPortugal1984; @onezimeNewGeodynamicInterpretation2003], located in the north sector of the region, and characterized by schists, greywackes and quartzites; and 2) the Algarve basin in the south, which is characterized by limestones, dolomites and marls [@terrinhaBaciaAlgarveEstratigrafia2006]. It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules/beds can be found [@manuppellaNoticiaExplicativaFolha1987; @manuppellaNoticiaExplicativaFolha2007; @ribeiro_evolucao_2005; @rochaNoticiaExplicativaFolha1979; @rochaNoticiaExplicativaFolha1983; @terrinhaBaciaAlgarveEstratigrafia2006].

@@

We employed both macroscopic and petrographic methods to achieve a more comprehensive characterization of the materials, aiming to offset the inherent limitations of each method [@luedtkeArchaeologistGuideChert1992].

Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustilloMacroscopicClassificationFlint2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=4458). The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins. Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustilloMacroscopicClassificationFlint2009; @tarrinoMateriasPrimasLiticas2013]. Therefore, we implemented an initial phase of macroscopic analysis, subsequently augmented by petrographic study to diminish the subjectivity of the analysis and enhance the material characterization with petrographic data. The synergistic application of these two methods has been previously validated in the characterization of Algarve region cherts, following recent geological survey efforts [@belmiroCreatingFramesReference2023]. Hence, an extensive macroscopic and petrographic collection was already available for comparative analysis with archaeological materials, employing consistent methodologies. This reference collection is the Lusolit lithotheque, hosted physically at ICArEHB (University of Algarve, Faro) and digitally at www.lusolit.icarehb.com, and aided in the characterization and source attribution of cherts.

@@

The macroscopic analysis of the archaeological materials was divided into two steps. The first step included a preliminary characterization using a hand lens of 10x magnification. This step allowed us to create types and subtypes based on similar macroscopic characteristics frequently acknowledged to be useful to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtkeArchaeologistGuideChert1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1). In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types - the data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1). Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis was simplified and focused on collecting data related to weight, cortex, and alterations. This first step was applied to a total of 4458 artefacts: a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 7 to 4) and 831 chert artefacts from the Solutrean sequence of the Shelter area (layers A, B, C; all samples without an attributed layer on the database were excluded from the analysis).

@@

```{r}
#| label: fig-local-types
#| fig-cap: "Variability of local chert types and main petrographic characteristics. a) Type 1 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope of white patina and fractures; a2) microscopic view of thin section (XPL), with visible fibrous chalcedony composition and without allochems. b) Type 3 macroscopic and microscopic views: b1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; b2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides. c-d) Type 2 macroscopic and microscopic views of multicoloured (c) and single-colour (d) varieties. c1) macroscopic view with a stereomicroscope of the multicoloured bands with iron oxides and unidentifiable fossils; d1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; c2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; d2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides. e-f) Type 4 macroscopic and microscopic views: e) macroscopic view with a stereomicroscope of unidentifiable fossils and white splotches; f1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils within white and yellow splotches; f2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides; f3) microscopic view of thin section (PPL) with a concentration of fossils (possibly sponge spicules) and a fracture. g-h) Type 5 macroscopic and microscopic views: g) macroscopic view with a stereomicroscope; h1) macroscopic view with a stereomicroscope with visible unidentifiable fossils; h2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; h3) microscopic view of thin section (XPL) with small unidentifiable fossils, iron oxides and a possible ostracod replaced by fibrous chalcedony and macrocrystalline quartz."

@@

```{r}
#| label: fig-distance-types
#| fig-cap: "Variability of non-local types and main petrographic characteristics. a-b) Type 6 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope with a small degree of white patina and no visible fossils; b1) macroscopic view with a stereomicroscope of a thick alteration rind but translucent interior where some unidentifiable fossils are visible; a2) microscopic view of thin section (PPL) with a mudstone texture; b2) microscopic of thin section (XPL) with small unidentifiable fossils and a laminar microstructure. c-d) Type 7 macroscopic and microscopic views of the peloidal facies (c) and oolitic facies (d): c1) macroscopic view with a stereomicroscope of peloids and sponge spicules; d1) macroscopic view with a stereomicroscope of poorly preserved ooids; c2) microscopic view of thin section (PPL) with packstone texture composed mainly of peloids; d2) microscopic view of thin section (PPL) with packstone texture composed mainly of ooids. e-f) Type 8 macroscopic and microscopic views, with foraminifera represented by the white and black arrows: e) macroscopic view with a stereomicroscope with abundant unidentifiable fossils and foraminifera; f1) macroscopic view with a stereomicroscope with abundant unidentifiable fossils but no visible foraminifera; f2) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which are not visible macroscopically; f3) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which not are visible macroscopically. g) Type 9 macroscopic and microscopic views: g1) macroscopic view with a stereomicroscope of unidentifiable fossils and two different fabrics; g2) microscopic view of thin section (XPL) with packstone texture composed of concentrations of unidentifiable fossils and opaques which create a banded microstructure. h) Type 10 macroscopic and microscopic views: h1) macroscopic view with a stereomicroscope of ooids and fractures filled with chalcedony; h2) microscopic view of thin section (PPL) with packstone texture composed of ooids with different levels of preservation."

@@

Type 10 (@fig-distance-types h) is an oolitic packstone, massive micro-cryptocrystalline micrite/quartz from a marine environment and possibly a high-energy, shallow depositional environment, due to the type, sorting, and preservation conditions of the ooids. It represents \~1% of the overall Terrace cherts and less than 0.1% of the chert in the Shelter. Its heterogeneous structure has two types of fabrics with different macroscopic characteristics: brown (opaque) and black (sub-translucent) fabric (@fig-distance-types h1). The sample chosen for the thin section is composed of 35% micro-cryptocrystalline quartz, 53% micrite, 9% macrocrystalline quartz, and accessory fibrous chalcedony (filling fractures) and mica; the percentages of quartz and micrite might be related to the fabrics present in the artefacts. Allochems are common iron oxides and very frequent ooids. Bioclasts are rare, poorly preserved, and replaced by quartz. The porosity is intraparticle and vug (10%). The ooids are distributed homogeneously across the sample (although macroscopically visible in the brown fabric) and are highly abundant and concentrated. They are poorly sorted, varying between 500 and 20 μm (@fig-distance-types h2), show round-to-elliptical shapes and their preservation is variable: some ooids show a poorly preserved micritic structure, while others show concentric laminae structures around a round nucleus [@flugelMicrofaciesCarbonateRocks2010]. This chert shows no similarities to any consulted reference material. Despite the existence of oolitic limestones in the Algarve region correspondent to Middle and Upper Jurassic formations and often described as whiteish (Rocha et al., 1979), the presence of oolitic chert in the region is not recognized in geological and archaeological literature [@cardosoSobrePresencaLaminas2018; @rochaNoticiaExplicativaFolha1979; @verissimo_jazidas_2004] and the visit of outcrops mentioned in the geological cartography did not allow their identification.

@@

```{r}
#| label: fig-small-types
#| fig-cap: "Macroscopic view with a stereomicroscope of non-local Type 11 and TL cherts. Type 11 – detail of banded patterns composed of parallel different fabrics; TL01 – detail of the jasper and fractures filled with chalcedony; TL02 – detail of the chalcedony with unidentified orange inclusions; TL03 – detail of two round inclusions, possibly iron oxides; TL04 - detail of massive chert with fractures filled with chalcedony; TL05 – detail of the banded pattern within the grey mudstone texture; TL06 - detail of the small, finely sorted ooids; TL07 – detail of the dendritic laminations, possibly composed of iron oxides; TL08 – detail of the mudstone structure, without the presence of allochems; TL09a/b – detail of the mossy, orange and red inclusions within the chalcedony; TL10 – detail of the speckling and laminated patterns within the mudstone structure; TL11 – detail of the banding within the mudstone structure; TL12 – detail of the poorly preserved ooids, only visible under the stereomicroscope; TL13 – detail of the bands and laminations within the mudstone texture fabric; TL14 – detail of an unidentified benthic foraminifera; TL15 – detail of the interface between the massive brown fabric and splotches; TL16 - detail of the of the massive structure with unidentifiable inclusions."

@@

Jasper is known in the South Portuguese Zone area [@oliveiraCartaGeologicaPortugal1984], from Southern/Central Portugal to Spain, north of the Algarve (from Aljustrel to Beja; @fig-map), suggesting that the source of TL01 may be from south/central Portugal at a minimum distance of 100 km north and northeast from Vale Boi (as the crow flies).

@@

In the Gravettian occupations (levels 6 and 7) and at the start of the Proto-Solutrean (level 5) local cherts make up \~45-60%, while the combined non-local cherts make up \~65-40%. Level 6 shows the highest percentages of all non-local chert use, with specific types like Type 6 representing \~30%, although when considering the sum of weight, the percentages reduce, and non-local chert represents \~50% of total chert. The percentages of local chert increase in the following occupations. In the Proto-Solutrean (levels 4E and 4E/5) and Solutrean (levels 4, 4C, and 4D) local cherts make up more than \~70% of the sample and 80% when considering the sum of weight, while non-local cherts decrease significantly and represent percentages lower than 30%. Despite the increase of local cherts during the Proto-Solutrean and Solutrean, there are still significant differences between the local types present in the two technocomplexes. During the top Proto-Solutrean occupations the grey/white cherts (Type 4) are the most present chert with percentages between 35-20%, while in the Solutrean levels, the yellow/brown cherts (Type 2) are present in the highest percentage, at \~30%.

```{r}
#| label: fig-type-per-level
#| fig-cap: Terrace area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples. Only percentages superior to the 5% threshold are shown in the figure. Type 1 to 5 (yellow and red colours) represent local chalcedony and cherts. Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colours). Type 8, 10 and 11 represent non-local cherts with non-identified provenience (blue and green colours).
#| fig-width: 10

@@

When plotting the weight, it is apparent that most of the samples are located under the 10 gr mark, with median and bottom quartiles often falling under the 5 gr line (@fig-weight). Local types, such as Type 2, Type 4, and Type 5, although mostly represented by artefacts with a medium mass (\~5-10 gr), show the heaviest and highest number of heavy artefacts. In comparison, other types, although showing some variation in the weight of artefacts, show smaller artefacts and their means and medians are frequently closer to the bottom quartile. As seen in (@fig-weight), non-local types show larger weights from level 5 to level 7.

```{r}
#| label: fig-weight
#| fig-cap: "Terrace boxplots of chert type weight (gr) per level. Diamond shape within the boxplots represents weight mean. Type 1 to 5 (yellow and red colours) represent local chalcedony and cherts. Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colours). Type 8, 10 and 11 represent non-local cherts with non-identified provenience (blue and green colours)."
#| #| fig-width: 10

@@

```{r}
#| label: fig-analysis_shelter
#| fig-cap: "a-b) Shelter area chalcedony and chert type frequency (a) and weight sum (b) per level, without unidentifiable samples. Only percentages superior to the 5% threshold are shown in the figure. c): Shelter boxplots of chert type weight (gr) per level. Type 1 to 5 (yellow and red colours) represent local chalcedony and cherts. Type 6, 7 and 9 represent non-local cherts with identified probable provenience (brown colours). Type 8, 10 and 11 represent non-local cherts with non-identified provenience (blue and green colours)."
#| #| fig-width: 10

@@

A similar case is observed for the jaspers located in the SPZ [@oliveiraCartaGeologicaPortugal1984], located at a minimum of 100 km distance from Vale Boi, as the crow flies. Despite being a source located at closer distances when compared to the cherts of Central Portugal or from Southwestern Spain, the jasper occurs in trace amounts (n=3) and only in the Gravettian layers (layer 6). Occurring North and Northeast, the limited amount of jasper in the assemblage may either represent a limited use of these resources in detriment of the cherts from the Lisbon and Rio Maior area, or suggest that they were used and knapped elsewhere, but not transported, either by the same groups or through exchanges by different groups.
